---
title: "STAT545_hw04_JasmineLib"
output: github_document
---
### STAT 545 Homework 4

Data Reshaping Prompts (and relationship to aggregation)

```{r}
suppressPackageStartupMessages(library(tidyverse)) 
suppressPackageStartupMessages(library(gapminder))
```

```{r}
test = gapminder %>% 
  select (year, lifeExp, country) %>% 
  filter (country == "China" | country == "India" | country == "Japan") %>% 
   spread(key = "country", value = "lifeExp")
test

class(test)

test  %>% 
  knitr::kable()
  

test %>%
  ggplot (aes(China, India)) + 
  geom_point(aes(colour = year)) + 
  theme_classic() +
  ggtitle("Life Expectancy in China vs. India") +
  labs(x = "China Life Expectancy (Years)", y = "India Life Expectancy (Years)") +
  scale_y_continuous(breaks=seq(from = 25, to =75, by=5)) +
  scale_x_continuous(breaks=seq(30,70, by=5)) +
  geom_abline(mapping = NULL, data = NULL, slope = 1,
  na.rm = FALSE) +
  theme(aspect.ratio=1) 
```

Join Prompts (join, merge, look up)

Activity #1

Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:


One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.
One row per continent, a continent variable and one or more variables with extra info, such as northern versus southern hemisphere.
```{r}
#Obtained data from: https://www.gapminder.org/data/
numHIV = read_csv(file = "people_living_with_hiv_number_all_ages.csv")

gapminder_sel = gapminder %>% 
  filter (year == "2007") %>% 
  select(country = "country", pop, gdpPercap, continent) #select the population and gdpPercap data for the year 2007. 
  #filter(country =="Canada" | country == "Japan" | country =="India")


gapminder_sel 

numHIV_sel =numHIV %>% 
  select(geo, "2007") #rename the "geo" column to country to make joining easier, and select data for the year 2007. 
  #filter(geo =="Canada" | geo == "Japan" | geo =="India")
 
class(gapminder$country)
class(numHIV$geo)

gapminderHIV = left_join(gapminder_sel, numHIV_sel, by= c("country" = "geo")) #left join columns, join by comparing country and geo columns. 

gapminderHIV %>% 
  filter(continent =="Americas") %>% 
  mutate(HIVper10000 = `2007`/pop*10000) %>% 
  ggplot(aes(country, HIVper10000)) + 
  geom_point(na.rm = TRUE) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Country", y = "HIV frequency per 10 000 people")

```

